Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2019-04-03 submission/manuscript.Rmd
Edits

# Fecal short chain fatty acids are not predictive of colonic tumor status and cannot be predicted based on bacterial community structure

@@

Colonic bacterial populations are thought to have a role in the development of colorectal cancer with some protecting against inflammation and others exacerbating inflammation. Short chain fatty acids (SCFAs), including butyrate, have been shown to have anti-inflammatory properties and are produced in large quantities by colonic bacteria which produce SCFAs by fermenting fiber. We assessed whether there was an association between fecal SCFA concentrations and the presence of colonic adenomas or carcinomas in a cohort of individuals using 16S rRNA gene and metagenomic shotgun sequence data. We measured the fecal concentrations of acetate, propionate, and butyrate within the cohort and found that there were no significant associations between SCFA concentration and disease status. When we incorporated these concentrations into random forest classification models trained to differentiate between people with normal colons and those with adenomas or carcinomas, we found that they did not significantly improve the ability of 16S rRNA gene or metagenomic gene sequence-based models to classify individuals. Finally, we generated random forest regression models trained to predict the concentration of each SCFA based on 16S rRNA gene or metagenomic gene sequence data from the same samples. These models performed poorly and were able to explain at most 14% of the observed variation in the SCFA concentrations. These results support the broader epidemiological data that questions the value of fiber consumption for reducing the risks of colorectal cancer. Although other bacterial metabolites may serve as biomarkers to detect adenomas or carcinomas, fecal SCFA concentrations have limited predictive power.

@@

Considering colorectal cancer is the third leading cancer-related cause of death within the United States, there is a great need to detect colorectal tumors early without invasive colonoscopy procedures and to prevent the formation of tumors. Short chain fatty acids (SCFAs) are often used as a surrogate for measuring gut health and for being anti-carcinogenic because of their anti-inflammatory properties. We evaluated the fecal SCFA concentration of a cohort of individuals with varying colonic tumor burden who were previously analyzed to identify microbiome-based biomarkers of tumors. We were unable to find an association between SCFA concentration and tumor burden or use SCFAs to improve our microbiome-based models of classifying people based on their tumor status. Furthermore, we were unable to find an association between the fecal community structure and SCFA concentrations. These data indicate that there is no conclusive link between the gut microbiome, SCFAs, and tumor burden.

@@

Colorectal cancer is the third leading cancer-related cause of death within the United States [@Siegel2016]. Less than 10% of cases can be attributed to genetic risk factors [@Fearon1990]. This leaves a significant role for environmental and behavioral factors and diet [@FlissIsakov2017; @Lee2015]. Colorectal cancer is thought to be initiated by a series of mutations that accumulate as the mutated cells begin to proliferate leading to adenomatous lesions, which are succeeded by carcinomas [@Fearon1990]. Throughout this progression, there are ample opportunities for bacterial populations to have a role as bacteria are known to cause mutations, induce inflammation, and accelerate tumorigenesis [@Zackular2013; @DeStefanoShields2016; @Tomkovich2017]. Additional cross sectional studies in humans have identified microbiome-based biomarkers of disease [@Sze2018]. These studies suggest that in some cases, it is the loss of bacterial populations that produce short-chain fatty acids (SCFAs) that results in increased inflammation and tumorigenesis.

Many microbiome studies use the concentrations of SCFAs and the presence of 16S rRNA gene sequences from organisms and the genes involved in producing them as a biomarker of a healthy microbiota [@Sanna2019; @Meisel2016]. SCFAs have anti-inflammatory and anti-proliferative activities [@OKeefe2016]. Direct supplementation of SCFAs or feeding of fiber caused an overall reduction in tumor burden in mouse models of colorectal cancer [@Bishehsari2018]. These results suggest that supplementation with fiber, which many colonic bacteria ferment to produce SCFAs may confer beneficial effects against colorectal cancer. Regardless, there is a lack of evidence that increasing SCFA concentrations can protect against colorectal cancer in humans. Case-control studies that have investigated possible associations between SCFAs and colon tumor status have been plagued by relatively small numbers of subjects, but have reported increased total and relative fecal acetate levels and decreased relative fecal butyrate concentrations in subjects with colonic lesions [@Weaver1988]. In randomized controlled trials fiber supplementation has been inconsistently associated with protection against tumor formation and recurrence [@Yao2017; @Gianfredi2018]. These findings temper enthusiasm for treatments that aim to use SCFAs as biomarkers or protection against tumorigenesis.

@@

**Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas.** We previously found that binning 16S rRNA gene sequence data into operational taxonomic units (OTUs) based on 97% similarity or into genera enabled us to classify individuals as having adenomas or carcinomas using random forest machine learning models [@Baxter2016; @Sze2018]. We repeated that analysis but added the concentration of the SCFAs as possible features to train the models (Figure S1). Models trained using SCFAs to classify individuals as having adenomas or carcinomas rather than normal colons had median areas under the receiver operator characteristic curve (AUROC) that were significantly greater than 0.5 (P~adenoma~<0.001 and P~carcinoma~<0.001). However, the AUROC values to detect the presence of adenomas or carcinomas were only `r random %>% filter(class=="adenoma") %>% pull(median)` and `r random %>% filter(class=="cancer") %>% pull(median)`, respectively, indicating that SCFAs had poor predictive power on their own (Figure 2A). When we trained the models with the SCFAs concentrations and OTU or genus-level relative abundances the AUROC values were not significantly different from the same models trained without the SCFA concentrations (P>`r min_p_compare_tax`; Figure 2A). These data demonstrate that knowledge of the SCFA profile from a patient's fecal sample did not improve the ability to diagnose a colonic lesion.

@@

**Knowledge of microbial community structure does not predict SCFA concentrations.** We next asked whether the fecal community structure was predictive of fecal SCFA concentrations, regardless of a person's diagnosis. We trained random forest regression models using 16S rRNA gene sequence data binned into OTUs and genera to predict the concentration of the SCFAs (Figure S2). The largest R^2^ between the observed SCFA concentrations and the modeled concentrations was `r top_16S$median_Rsquared`, which was observed when using `r convert[top_16S$microbiome]` data to predict `r top_16S$class` concentrations (Figure 2B). We also used a smaller dataset of shotgun metagenomic sequencing data generated from a subset of our cohort (N~normal~=`r n_s %>% filter(dx == "normal") %>% pull(n)`, N~adenoma~=`r n_s %>% filter(dx == "adenoma") %>% pull(n)`, and N~cancer~=`r n_s %>% filter(dx == "cancer") %>% pull(n)`) [@Hannigan2017]. We binned genes extracted from the assembled metagenomes into operational protein families (OPFs) or KEGG categories and trained random forest regression models using metagenomic sequence data to predict the concentration of the SCFAs (Figure S2). Similar to the analysis using 16S rRNA gene sequence data, the metagenomic data was not predictive of SCFA concentration; the largest amount of variation that the models could explain was `r 100*top_metagenomics$median_Rsquared`%, which was observed when using `r convert[top_metagenomics$microbiome]` data to predict `r top_metagenomics$class` concentrations (Figure 2B). Because of the limited number of samples that we were able to generate metagenomic sequence data from, we used our 16S rRNA gene sequence data to impute metagenomes that were binned into metabolic pathways or KEGG categories using PICRUSt (Figure S2). SCFA concentrations could not be predicted based on the imputed metagenomic data. The largest amount of variation that the models could explain was `r 100*top_picrust$median_Rsquared`%, which was observed when using `r convert[top_picrust$microbiome]` data to predict `r top_picrust$class` concentrations (Figure 2B). The inability to model SCFA concentrations from microbiome data indicates that the knowledge of the abundance of organisms and their genes was insufficient to predict SCFA concentrations.

**Conclusion.** Our data indicate that fecal SCFA concentrations are not associated with the presence of adenomas or carcinomas and that they provide weak predictive power to improve the ability to diagnose someone with one of these lesions. Furthermore, knowledge of the taxonomic and genetic structure of gut microbiota was not predictive of SCFA concentrations. These results complement existing literature that suggest that fiber consumption and the production of SCFAs are unable to prevent the risk of developing colonic tumors. It is important to note that our analysis concerned fecal SCFA concentrations and microbiome characterization and that observations along the mucosa near the site of lesions may provide a stronger association. Regardless, given the growing literature in this area, it is unlikely that SCFAs are the primary mechanism that limits tumorigenesis. This may be a cautionary result to temper enthusiasm for SCFAs as a biomarker of gut health more generally. Going forward it is critical to develop additional hypotheses for how the microbiome and host interact to drive tumorigenesis so that we can better understand tumorigenesis and identify biomarkers that will allow early detection of lesions.

>>>>

# Fecal short-chain fatty acids are not predictive of colonic tumor status and cannot be predicted based on bacterial community structure

@@

Colonic bacterial populations are thought to have a role in the development of colorectal cancer with some protecting against inflammation and others exacerbating inflammation. Short-chain fatty acids (SCFAs), including butyrate, have been shown to have anti-inflammatory properties and are produced in large quantities by colonic bacteria which produce SCFAs by fermenting fiber. We assessed whether there was an association between fecal SCFA concentrations and the presence of colonic adenomas or carcinomas in a cohort of individuals using 16S rRNA gene and metagenomic shotgun sequence data. We measured the fecal concentrations of acetate, propionate, and butyrate within the cohort and found that there were no significant associations between SCFA concentration and tumor status. When we incorporated these concentrations into random forest classification models trained to differentiate between people with normal colons and those with adenomas or carcinomas, we found that they did not significantly improve the ability of 16S rRNA gene or metagenomic gene sequence-based models to classify individuals. Finally, we generated random forest regression models trained to predict the concentration of each SCFA based on 16S rRNA gene or metagenomic gene sequence data from the same samples. These models performed poorly and were able to explain at most 14% of the observed variation in the SCFA concentrations. These results support the broader epidemiological data that questions the value of fiber consumption for reducing the risks of colorectal cancer. Although other bacterial metabolites may serve as biomarkers to detect adenomas or carcinomas, fecal SCFA concentrations have limited predictive power.

@@

Considering colorectal cancer is the third leading cancer-related cause of death within the United States, there is a great need to detect colorectal tumors early without invasive colonoscopy procedures and to prevent the formation of tumors. Short-chain fatty acids (SCFAs) are often used as a surrogate for measuring gut health and for being anti-carcinogenic because of their anti-inflammatory properties. We evaluated the fecal SCFA concentration of a cohort of individuals with varying colonic tumor burden who were previously analyzed to identify microbiome-based biomarkers of tumors. We were unable to find an association between SCFA concentration and tumor burden or use SCFAs to improve our microbiome-based models of classifying people based on their tumor status. Furthermore, we were unable to find an association between the fecal community structure and SCFA concentrations. These data indicate that there is no conclusive link between the gut microbiome, SCFAs, and tumor burden.

@@

Colorectal cancer is the third leading cancer-related cause of death within the United States [@Siegel2016]. Less than 10% of cases can be attributed to genetic risk factors [@Fearon1990]. This leaves a significant role for environmental, behavioral, and dietary factors [@FlissIsakov2017; @Lee2015]. Colorectal cancer is thought to be initiated by a series of mutations that accumulate as the mutated cells begin to proliferate leading to adenomatous lesions, which are succeeded by carcinomas [@Fearon1990]. Throughout this progression, there are ample opportunities for bacterial populations to have a role as bacteria are known to cause mutations, induce inflammation, and accelerate tumorigenesis [@Zackular2013; @DeStefanoShields2016; @Tomkovich2017]. Additional cross sectional studies in humans have identified microbiome-based biomarkers of disease [@Sze2018]. These studies suggest that in some cases, it is the loss of bacterial populations that produce short-chain fatty acids (SCFAs) that results in increased inflammation and tumorigenesis.

Many microbiome studies use the concentrations of SCFAs and the presence of 16S rRNA gene sequences from organisms and the genes involved in producing them as a biomarker of a healthy microbiota [@Sanna2019; @Meisel2016]. SCFAs have anti-inflammatory and anti-proliferative activities [@OKeefe2016]. Direct supplementation of SCFAs or feeding of fiber caused an overall reduction in tumor burden in mouse models of colorectal cancer [@Bishehsari2018]. These results suggest that supplementation with fiber, which many colonic bacteria ferment to produce SCFAs may confer beneficial effects against colorectal cancer. Regardless, there is a lack of evidence that increasing SCFA concentrations can protect against colorectal cancer in humans. Case-control studies that have investigated possible associations between SCFAs and colon tumor status have been plagued by relatively small numbers of subjects, but have reported increased total and relative fecal acetate levels and decreased relative fecal butyrate concentrations in subjects with colonic lesions [@Weaver1988]. In randomized controlled trials fiber supplementation has been inconsistently associated with protection against tumor formation and recurrence [@Yao2017; @Gianfredi2018]. These findings temper enthusiasm for treatments that target the production of SCFAs or for using them as biomarkers for protection against tumorigenesis.

@@

**Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas.** We previously found that binning 16S rRNA gene sequence data into operational taxonomic units (OTUs) based on 97% similarity or into genera enabled us to classify individuals as having adenomas or carcinomas using random forest machine learning models [@Baxter2016; @Sze2018]. We repeated that analysis but added the concentration of the SCFAs as possible features to train the models (Figure S1). Models trained using SCFAs to classify individuals as having adenomas or carcinomas rather than normal colons had median areas under the receiver operator characteristic curve (AUROC) that were significantly greater than 0.5 (P~adenoma~<0.001 and P~carcinoma~<0.001). However, the AUROC values to detect the presence of adenomas or carcinomas were only `r random %>% filter(class=="adenoma") %>% pull(median)` and `r random %>% filter(class=="cancer") %>% pull(median)`, respectively, indicating that SCFAs had poor predictive power on their own (Figure 2A). When we trained the models with the SCFAs concentrations and OTU or genus-level relative abundances the AUROC values were not significantly different from the same models trained without the SCFA concentrations (P>`r min_p_compare_tax`; Figure 2A). These data demonstrate that knowledge of the SCFA profile from a subject's fecal sample did not improve the ability to diagnose a colonic lesion.

@@

**Knowledge of microbial community structure does not predict SCFA concentrations.** We next asked whether the fecal community structure was predictive of fecal SCFA concentrations, regardless of a person's diagnosis. We trained random forest regression models using 16S rRNA gene sequence data binned into OTUs and genera to predict the concentration of the SCFAs (Figure S2). The largest R^2^ between the observed SCFA concentrations and the modeled concentrations was `r top_16S$median_Rsquared`, which was observed when using `r convert[top_16S$microbiome]` data to predict `r top_16S$class` concentrations (Figure 2B). We also used a smaller dataset of shotgun metagenomic sequencing data generated from a subset of our cohort (N~normal~=`r n_s %>% filter(dx == "normal") %>% pull(n)`, N~adenoma~=`r n_s %>% filter(dx == "adenoma") %>% pull(n)`, and N~cancer~=`r n_s %>% filter(dx == "cancer") %>% pull(n)`) [@Hannigan2017]. We binned genes extracted from the assembled metagenomes into operational protein families (OPFs) or KEGG categories and trained random forest regression models using metagenomic sequence data to predict the concentration of the SCFAs (Figure S2). Similar to the analysis using 16S rRNA gene sequence data, the metagenomic data was not predictive of SCFA concentration. The largest R^2^ was `r top_metagenomics$median_Rsquared`, which was observed when using `r convert[top_metagenomics$microbiome]` data to predict `r top_metagenomics$class` concentrations (Figure 2B). Because of the limited number of samples that we were able to generate metagenomic sequence data from, we used our 16S rRNA gene sequence data to impute metagenomes that were binned into metabolic pathways or KEGG categories using PICRUSt (Figure S2). SCFA concentrations could not be predicted based on the imputed metagenomic data. The largest R^2^ was `r top_picrust$median_Rsquared`, which was observed when using `r convert[top_picrust$microbiome]` data to predict `r top_picrust$class` concentrations (Figure 2B). The inability to model SCFA concentrations from microbiome data indicates that the knowledge of the abundance of organisms and their genes was insufficient to predict SCFA concentrations.

**Conclusion.** Our data indicate that fecal SCFA concentrations are not associated with the presence of adenomas or carcinomas and that they provide weak predictive power to improve the ability to diagnose someone with one of these lesions. Furthermore, knowledge of the taxonomic and genetic structure of gut microbiota was not predictive of SCFA concentrations. These results complement existing literature that suggest that fiber consumption and the production of SCFAs are unable to prevent the risk of developing colonic tumors. It is important to note that our analysis was based on characterizations of SCFA and microbiome profiles using fecal samples and that observations along the mucosa near the site of lesions may provide a stronger association. This may be a cautionary result to temper enthusiasm for SCFAs as a biomarker of gut health more generally. Going forward it is critical to develop additional hypotheses for how the microbiome and host interact to drive tumorigenesis so that we can better understand tumorigenesis and identify biomarkers that will allow early detection of lesions.

### 2019-04-03 submission/manuscript.tex
Edits

\hypertarget{fecal-short-chain-fatty-acids-are-not-predictive-of-colonic-tumor-status-and-cannot-be-predicted-based-on-bacterial-community-structure}{%
\section{Fecal short chain fatty acids are not predictive of colonic
tumor status and cannot be predicted based on bacterial community
structure}\label{fecal-short-chain-fatty-acids-are-not-predictive-of-colonic-tumor-status-and-cannot-be-predicted-based-on-bacterial-community-structure}}

@@

Colonic bacterial populations are thought to have a role in the
development of colorectal cancer with some protecting against
inflammation and others exacerbating inflammation. Short chain fatty
acids (SCFAs), including butyrate, have been shown to have
anti-inflammatory properties and are produced in large quantities by
colonic bacteria which produce SCFAs by fermenting fiber. We assessed
whether there was an association between fecal SCFA concentrations and
the presence of colonic adenomas or carcinomas in a cohort of
individuals using 16S rRNA gene and metagenomic shotgun sequence data.
We measured the fecal concentrations of acetate, propionate, and
butyrate within the cohort and found that there were no significant
associations between SCFA concentration and disease status. When we
incorporated these concentrations into random forest classification
models trained to differentiate between people with normal colons and
those with adenomas or carcinomas, we found that they did not
significantly improve the ability of 16S rRNA gene or metagenomic gene
sequence-based models to classify individuals. Finally, we generated
random forest regression models trained to predict the concentration of
each SCFA based on 16S rRNA gene or metagenomic gene sequence data from
the same samples. These models performed poorly and were able to explain
at most 14\% of the observed variation in the SCFA concentrations. These
results support the broader epidemiological data that questions the
value of fiber consumption for reducing the risks of colorectal cancer.
Although other bacterial metabolites may serve as biomarkers to detect
adenomas or carcinomas, fecal SCFA concentrations have limited
predictive power.

@@

Colorectal cancer is the third leading cancer-related cause of death
within the United States (1). Less than 10\% of cases can be attributed
to genetic risk factors (2). This leaves a significant role for
environmental and behavioral factors and diet (3, 4). Colorectal cancer
is thought to be initiated by a series of mutations that accumulate as
the mutated cells begin to proliferate leading to adenomatous lesions,
which are succeeded by carcinomas (2). Throughout this progression,
there are ample opportunities for bacterial populations to have a role
as bacteria are known to cause mutations, induce inflammation, and
accelerate tumorigenesis (5--7). Additional cross sectional studies in
humans have identified microbiome-based biomarkers of disease (8). These
studies suggest that in some cases, it is the loss of bacterial
populations that produce short-chain fatty acids (SCFAs) that results in
increased inflammation and tumorigenesis.

Many microbiome studies use the concentrations of SCFAs and the presence
of 16S rRNA gene sequences from organisms and the genes involved in
producing them as a biomarker of a healthy microbiota (9, 10). SCFAs
have anti-inflammatory and anti-proliferative activities (11). Direct
supplementation of SCFAs or feeding of fiber caused an overall reduction
in tumor burden in mouse models of colorectal cancer (12). These results
suggest that supplementation with fiber, which many colonic bacteria
ferment to produce SCFAs may confer beneficial effects against
colorectal cancer. Regardless, there is a lack of evidence that
increasing SCFA concentrations can protect against colorectal cancer in
humans. Case-control studies that have investigated possible
associations between SCFAs and colon tumor status have been plagued by
relatively small numbers of subjects, but have reported increased total
and relative fecal acetate levels and decreased relative fecal butyrate
concentrations in subjects with colonic lesions (13). In randomized
controlled trials fiber supplementation has been inconsistently
associated with protection against tumor formation and recurrence (14,
15). These findings temper enthusiasm for treatments that aim to use
SCFAs as biomarkers or protection against tumorigenesis.

@@

\textbf{Combining SCFA and microbiome data does not improve the ability
to diagnose individual as having adenomas or carcinomas.} We previously
found that binning 16S rRNA gene sequence data into operational
taxonomic units (OTUs) based on 97\% similarity or into genera enabled
us to classify individuals as having adenomas or carcinomas using random
forest machine learning models (8, 16). We repeated that analysis but
added the concentration of the SCFAs as possible features to train the
models (Figure S1). Models trained using SCFAs to classify individuals
as having adenomas or carcinomas rather than normal colons had median
areas under the receiver operator characteristic curve (AUROC) that were
significantly greater than 0.5 (P\textsubscript{adenoma}\textless{}0.001
and P\textsubscript{carcinoma}\textless{}0.001). However, the AUROC
values to detect the presence of adenomas or carcinomas were only 0.54
and 0.55, respectively, indicating that SCFAs had poor predictive power
on their own (Figure 2A). When we trained the models with the SCFAs
concentrations and OTU or genus-level relative abundances the AUROC
values were not significantly different from the same models trained
without the SCFA concentrations (P\textgreater{}0.21; Figure 2A). These
data demonstrate that knowledge of the SCFA profile from a patient's
fecal sample did not improve the ability to diagnose a colonic lesion.

\textbf{Knowledge of microbial community structure does not predict SCFA
concentrations.} We next asked whether the fecal community structure was
predictive of fecal SCFA concentrations, regardless of a person's
diagnosis. We trained random forest regression models using 16S rRNA
gene sequence data binned into OTUs and genera to predict the
concentration of the SCFAs (Figure S2). The largest R\textsuperscript{2}
between the observed SCFA concentrations and the modeled concentrations
was 0.14, which was observed when using genus data to predict butyrate
concentrations (Figure 2B). We also used a smaller dataset of shotgun
metagenomic sequencing data generated from a subset of our cohort
(N\textsubscript{normal}=27, N\textsubscript{adenoma}=25, and
N\textsubscript{cancer}=26) (18). We binned genes extracted from the
assembled metagenomes into operational protein families (OPFs) or KEGG
categories and trained random forest regression models using metagenomic
sequence data to predict the concentration of the SCFAs (Figure S2).
Similar to the analysis using 16S rRNA gene sequence data, the
metagenomic data was not predictive of SCFA concentration; the largest
amount of variation that the models could explain was 5.45\%, which was
observed when using KEGG data to predict propionate concentrations
(Figure 2B). Because of the limited number of samples that we were able
to generate metagenomic sequence data from, we used our 16S rRNA gene
sequence data to impute metagenomes that were binned into metabolic
pathways or KEGG categories using PICRUSt (Figure S2). SCFA
concentrations could not be predicted based on the imputed metagenomic
data. The largest amount of variation that the models could explain was
9.67\%, which was observed when using KEGG data to predict butyrate
concentrations (Figure 2B). The inability to model SCFA concentrations
from microbiome data indicates that the knowledge of the abundance of
organisms and their genes was insufficient to predict SCFA
concentrations.

\textbf{Conclusion.} Our data indicate that fecal SCFA concentrations
are not associated with the presence of adenomas or carcinomas and that
they provide weak predictive power to improve the ability to diagnose
someone with one of these lesions. Furthermore, knowledge of the
taxonomic and genetic structure of gut microbiota was not predictive of
SCFA concentrations. These results complement existing literature that
suggest that fiber consumption and the production of SCFAs are unable to
prevent the risk of developing colonic tumors. It is important to note
that our analysis concerned fecal SCFA concentrations and microbiome
characterization and that observations along the mucosa near the site of
lesions may provide a stronger association. Regardless, given the
growing literature in this area, it is unlikely that SCFAs are the
primary mechanism that limits tumorigenesis. This may be a cautionary
result to temper enthusiasm for SCFAs as a biomarker of gut health more
generally. Going forward it is critical to develop additional hypotheses
for how the microbiome and host interact to drive tumorigenesis so that
we can better understand tumorigenesis and identify biomarkers that will
allow early detection of lesions.

>>>>

\hypertarget{fecal-short-chain-fatty-acids-are-not-predictive-of-colonic-tumor-status-and-cannot-be-predicted-based-on-bacterial-community-structure}{%
\section{Fecal short-chain fatty acids are not predictive of colonic
tumor status and cannot be predicted based on bacterial community
structure}\label{fecal-short-chain-fatty-acids-are-not-predictive-of-colonic-tumor-status-and-cannot-be-predicted-based-on-bacterial-community-structure}}

@@

Colonic bacterial populations are thought to have a role in the
development of colorectal cancer with some protecting against
inflammation and others exacerbating inflammation. Short-chain fatty
acids (SCFAs), including butyrate, have been shown to have
anti-inflammatory properties and are produced in large quantities by
colonic bacteria which produce SCFAs by fermenting fiber. We assessed
whether there was an association between fecal SCFA concentrations and
the presence of colonic adenomas or carcinomas in a cohort of
individuals using 16S rRNA gene and metagenomic shotgun sequence data.
We measured the fecal concentrations of acetate, propionate, and
butyrate within the cohort and found that there were no significant
associations between SCFA concentration and tumor status. When we
incorporated these concentrations into random forest classification
models trained to differentiate between people with normal colons and
those with adenomas or carcinomas, we found that they did not
significantly improve the ability of 16S rRNA gene or metagenomic gene
sequence-based models to classify individuals. Finally, we generated
random forest regression models trained to predict the concentration of
each SCFA based on 16S rRNA gene or metagenomic gene sequence data from
the same samples. These models performed poorly and were able to explain
at most 14\% of the observed variation in the SCFA concentrations. These
results support the broader epidemiological data that questions the
value of fiber consumption for reducing the risks of colorectal cancer.
Although other bacterial metabolites may serve as biomarkers to detect
adenomas or carcinomas, fecal SCFA concentrations have limited
predictive power.

@@

Colorectal cancer is the third leading cancer-related cause of death
within the United States (1). Less than 10\% of cases can be attributed
to genetic risk factors (2). This leaves a significant role for
environmental, behavioral, and dietary factors (3, 4). Colorectal cancer
is thought to be initiated by a series of mutations that accumulate as
the mutated cells begin to proliferate leading to adenomatous lesions,
which are succeeded by carcinomas (2). Throughout this progression,
there are ample opportunities for bacterial populations to have a role
as bacteria are known to cause mutations, induce inflammation, and
accelerate tumorigenesis (5--7). Additional cross sectional studies in
humans have identified microbiome-based biomarkers of disease (8). These
studies suggest that in some cases, it is the loss of bacterial
populations that produce short-chain fatty acids (SCFAs) that results in
increased inflammation and tumorigenesis.

Many microbiome studies use the concentrations of SCFAs and the presence
of 16S rRNA gene sequences from organisms and the genes involved in
producing them as a biomarker of a healthy microbiota (9, 10). SCFAs
have anti-inflammatory and anti-proliferative activities (11). Direct
supplementation of SCFAs or feeding of fiber caused an overall reduction
in tumor burden in mouse models of colorectal cancer (12). These results
suggest that supplementation with fiber, which many colonic bacteria
ferment to produce SCFAs may confer beneficial effects against
colorectal cancer. Regardless, there is a lack of evidence that
increasing SCFA concentrations can protect against colorectal cancer in
humans. Case-control studies that have investigated possible
associations between SCFAs and colon tumor status have been plagued by
relatively small numbers of subjects, but have reported increased total
and relative fecal acetate levels and decreased relative fecal butyrate
concentrations in subjects with colonic lesions (13). In randomized
controlled trials fiber supplementation has been inconsistently
associated with protection against tumor formation and recurrence (14,
15). These findings temper enthusiasm for treatments that target the
production of SCFAs or for using them as biomarkers for protection
against tumorigenesis.

@@

\textbf{Combining SCFA and microbiome data does not improve the ability
to diagnose individual as having adenomas or carcinomas.} We previously
found that binning 16S rRNA gene sequence data into operational
taxonomic units (OTUs) based on 97\% similarity or into genera enabled
us to classify individuals as having adenomas or carcinomas using random
forest machine learning models (8, 16). We repeated that analysis but
added the concentration of the SCFAs as possible features to train the
models (Figure S1). Models trained using SCFAs to classify individuals
as having adenomas or carcinomas rather than normal colons had median
areas under the receiver operator characteristic curve (AUROC) that were
significantly greater than 0.5 (P\textsubscript{adenoma}\textless{}0.001
and P\textsubscript{carcinoma}\textless{}0.001). However, the AUROC
values to detect the presence of adenomas or carcinomas were only 0.54
and 0.55, respectively, indicating that SCFAs had poor predictive power
on their own (Figure 2A). When we trained the models with the SCFAs
concentrations and OTU or genus-level relative abundances the AUROC
values were not significantly different from the same models trained
without the SCFA concentrations (P\textgreater{}0.21; Figure 2A). These
data demonstrate that knowledge of the SCFA profile from a subject's
fecal sample did not improve the ability to diagnose a colonic lesion.

\textbf{Knowledge of microbial community structure does not predict SCFA
concentrations.} We next asked whether the fecal community structure was
predictive of fecal SCFA concentrations, regardless of a person's
diagnosis. We trained random forest regression models using 16S rRNA
gene sequence data binned into OTUs and genera to predict the
concentration of the SCFAs (Figure S2). The largest R\textsuperscript{2}
between the observed SCFA concentrations and the modeled concentrations
was 0.14, which was observed when using genus data to predict butyrate
concentrations (Figure 2B). We also used a smaller dataset of shotgun
metagenomic sequencing data generated from a subset of our cohort
(N\textsubscript{normal}=27, N\textsubscript{adenoma}=25, and
N\textsubscript{cancer}=26) (18). We binned genes extracted from the
assembled metagenomes into operational protein families (OPFs) or KEGG
categories and trained random forest regression models using metagenomic
sequence data to predict the concentration of the SCFAs (Figure S2).
Similar to the analysis using 16S rRNA gene sequence data, the
metagenomic data was not predictive of SCFA concentration. The largest
R\textsuperscript{2} was 0.055, which was observed when using KEGG data
to predict propionate concentrations (Figure 2B). Because of the limited
number of samples that we were able to generate metagenomic sequence
data from, we used our 16S rRNA gene sequence data to impute metagenomes
that were binned into metabolic pathways or KEGG categories using
PICRUSt (Figure S2). SCFA concentrations could not be predicted based on
the imputed metagenomic data. The largest R\textsuperscript{2} was
0.097, which was observed when using KEGG data to predict butyrate
concentrations (Figure 2B). The inability to model SCFA concentrations
from microbiome data indicates that the knowledge of the abundance of
organisms and their genes was insufficient to predict SCFA
concentrations.

\textbf{Conclusion.} Our data indicate that fecal SCFA concentrations
are not associated with the presence of adenomas or carcinomas and that
they provide weak predictive power to improve the ability to diagnose
someone with one of these lesions. Furthermore, knowledge of the
taxonomic and genetic structure of gut microbiota was not predictive of
SCFA concentrations. These results complement existing literature that
suggest that fiber consumption and the production of SCFAs are unable to
prevent the risk of developing colonic tumors. It is important to note
that our analysis was based on characterizations of SCFA and microbiome
profiles using fecal samples and that observations along the mucosa near
the site of lesions may provide a stronger association. This may be a
cautionary result to temper enthusiasm for SCFAs as a biomarker of gut
health more generally. Going forward it is critical to develop
additional hypotheses for how the microbiome and host interact to drive
tumorigenesis so that we can better understand tumorigenesis and
identify biomarkers that will allow early detection of lesions.
